---
title: "Final Project STAT 331"
author: "Emma Turilli, John Ieng, Natalie Sakamoto, Gabby Apsay"
format: 
  html:
    code-tools: True
    toc: True
    embed-resources: True
    code-fold: True
editor: source
execute: 
  error: True
  message: False
  warning: False
bibliography: references.bib
---

```{r packages}
#| include: false

library(tidyverse)
library(knitr)
library(gganimate)
library(broom)
library(gifski)
library(DT)
library(gt)
library(kableExtra)
library(scales)
library(bibtex)
library(patchwork)
```

```{r seed}
#| include: false
set.seed(47984)
```

### Reproducibility

All code, raw data, and project files are available in our [GitHub Repository](https://github.com/gapsay/stat331-FinalProject/tree/main). Feel free to explore or replicate our analysis!

Life expectancy is a key indicator of a population's overall health. It can be attributed to various economic, social, and environmental factors. By exploring a nation's wealth, we would be able to analyze how economic prosperity impacts the longevity of the overall population.

# 1 Project Proposal + Data

This analysis utilizes the life expectancy and the gross domestic product (GDP) datasets sourced from Gapminder, a non-profit organization whose mission “is to fight devastating ignorance with a fact-based world view everyone could understand” [-@gapminderAbout]. Their site provides data sets collected from many reputable sources and interactive visualizations on important world topics. Additionally, we used a countries by continent dataset from Daina Bouquin posted on Kaggle [-@kaggleCountriesByContinent]. The dataset provides continent classifications for countries, allowing us conduct continent-specific analysis in our study.

We were interested in these data sets, as previous research has shown that GDP and life expectancy are related. A higher GDP per Capita allows for better access to essential resources and services that improve health and longevity, such as better diet, healthcare, and sanitation. Additionally, GDP serves as an indicator of economic development. Countries with higher GDPs often have more advanced infrastructure and social services, which further extend life expectancy [-@Jones_GDP].

```{r data}
lex <- read.csv("../stat331-FinalProject/lex.csv")
gdp_pcap <- read.csv("../stat331-FinalProject/gdp_pcap.csv")
continents <- read.csv("../stat331-FinalProject/country_by_continents.csv")
```

## 1.1 Data Cleaning

In the raw GDP dataset, some values used a “k” suffix for thousands (e.g., 10k for 10,000). We created a function to convert these abbreviations into numeric form, ensuring accurate comparisons and preventing data loss in analysis.

```{r}
clean_year <- function(str) {
  str <- case_when(
    str_detect(str, "k$") ~
      as.numeric(str_replace_all(str, "k$", "")) * 1000,
    .default = as.numeric(str)
  )
  return(str)
}
```

## 1.2 Pivoting Longer

The life expectancy data contains information about the life expectancy for 196 countries from the year 1800 to 2019, and predictive data from 2020 to 2100. It provides the life expectancy in years for each country within the set. For the period from 1800 to 1970, the data was sourced by Mattias Lindgren who assessed impacts from the biggest disasters in history in order to make rough estimates of life expectancy. Data for 1950-2019 was from the Global Burden of Disease Study 2019, which has 1950-2019 from the IHME. For 2020-2100, Gapminder used UN forecasts from the World Population Prospects 2022, [-@LEX].

**Life Expectancy Info from:** <https://www.gapminder.org/data/documentation/gd004>

```{r}
# pivot year columns to a single year and life expectancy column
lex <- lex |>
  pivot_longer(
    cols = X1800:X2100,
    names_to = "year",
    values_to = "life_expectancy"
  ) |>
  mutate(
    year = str_sub(year, start = 2),
    year = as.factor(year)
  )
```

The GDP data was obtained from the Madison Project Database (MPD) and Penn World Table (PWT). This data set contains information on gross domestic product (GDP) per person adjusted for differences in purchasing power in international dollars, and fixed 2017 prices. GDP per Capita measures the value of everything a country produces during a year, divided by the number of people. This set contains observations from 1800 to 2022. We transformed the data to have columns containing the country, year, and GDP of interest, [-@GDP].

**GDP Info from:** <https://www.gapminder.org/data/documentation/gd001/>

```{r}
# pivot year columns to a single year and gdp column
gdp_pcap <- gdp_pcap |>
  select(country, X1800:X2019) |>
  mutate(across(
    .cols = X1800:X2019,
    .fns = ~ clean_year(.x)
  )) |>
  pivot_longer(
    cols = X1800:X2019,
    names_to = "year",
    values_to = "gdp"
  ) |>
  mutate(
    year = str_sub(year, start = 2),
    year = as.factor(year),
    gdp = as.numeric(gdp)
  )
```

We transformed each of the individual year columns into one singular column so that the dataset would be easier to read. As a result, each observation consists of one country and year, with the corresponding life expectancy. The raw GDP data is similar to the life expectancy data in that each year has its own column. So we transformed the data in a similar way, making year its own column with its corresponding GDP. For accuracy, we are only using observations from 1800 to 2019, and none of the predicted data.

## 1.3 Joining Datasets

```{r}
# join both tables by country and year
gdp_lex <- lex |>
  full_join(
    gdp_pcap,
    join_by(
      country == country,
      year == year
    )
  ) |>
  na.omit()

# creating a data frame to add continent
country_mapping <- tibble(
  continent_name = c(
    "Burkina", "Burma (Myanmar)", "Congo", "Czechia",
    "Democratic Republic of Congo", "East Timor", "Hong Kong",
    "Ivory Coast", "Kyrgyzstan", "Laos", "Macedonia", "Micronesia",
    "Saint Kitts and Nevis", "Saint Lucia",
    "Saint Vincent and the Grenadines", "Slovakia", "Swaziland",
    "United Arab Emirates", "United Kingdom", "United States", "Palestine"
  ),
  gdp_name = c(
    "Burkina Faso", "Myanmar", "Congo, Rep.", "Czech Republic",
    "Congo, Dem. Rep.", "Timor-Leste", "Hong Kong, China",
    "Cote d'Ivoire", "Kyrgyz Republic", "Lao",
    "North Macedonia", "Micronesia, Fed. Sts.", "St. Kitts and Nevis",
    "St. Lucia", "St. Vincent and the Grenadines",
    "Slovak Republic", "Eswatini", "UAE", "UK", "USA", "Palestine"
  )
)

# altering continent to include Palestine
continents <- continents |>
  add_row(Continent = "Asia", Country = "Palestine") |>
  left_join(country_mapping, by = c("Country" = "continent_name")) |>
  mutate(combined = if_else(!is.na(gdp_name), gdp_name, Country))

# Join both datasets to assign continent value to all countries
gdp_lex <- left_join(gdp_lex, continents,
  by = c("country" = "combined")
) |>
  select(-country, -gdp_name) |>
  mutate(Continent = as.factor(Continent),
         year = as.integer(year), 
         log_gdp = log(gdp)) |>
  drop_na()
```

After cleaning up each data set, we had to join the two together by our observational unit, country. We hypothesize that as GDP increases, life expectancy will also begin to increase, as a higher GDP correlates to better infrastructure and more/better access to healthcare and medicine.

# 2 Linear Regressions

## 2.1 Data Visualization

```{r}
#| fig-width: 12
#| fig.height: 8

# Define color palette for continents
continent_colors <- c(
  "Africa" = "#1b9e77",
  "Asia" = "#d95f02",
  "Europe" = "#7570b3",
  "North America" = "#e7298a",
  "Oceania" = "#66a61e",
  "South America" = "#e6ab02"
)

gdp_lex_mean <- gdp_lex |>
  group_by(Country) |>
  summarize(
    avg_gdp = mean(log_gdp, na.rm = TRUE),
    avg_life_expectancy = mean(life_expectancy, na.rm = TRUE),
    Continent = first(Continent)
  ) |>
  mutate(Continent = as.factor(Continent))

ggplot(gdp_lex_mean, aes(x = avg_gdp, y = avg_life_expectancy, color = Continent)) +
  scale_color_manual(values = continent_colors) +
  geom_point() +
  scale_alpha(range = c(0.2, 0.9), guide = "none") +
  theme_minimal() +
  labs(
    title = "GDP per Capita vs Life Expectancy",
    subtitle = "Non-linear relationship with confidence intervals by continent",
    x = "GDP per Capita",
    y = "Life Expectancy (years)"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 24),
    axis.title = element_text(size = 18),
    plot.subtitle = element_text(size = 18),
    plot.caption = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14)
  )
```

```{r}
#| fig-width: 12
#| fig.height: 8
#| eval: false
gdp_lex <- gdp_lex |> 
   mutate(year = as.integer(year)) |> 
   mutate(year = 1800 + (year - min(year))) |> 
   drop_na()

ggplot(gdp_lex, aes(log_gdp, life_expectancy, color = Continent)) +
  geom_point(alpha = 0.7, size = 2.5, show.legend = FALSE) +
  scale_color_manual(values = continent_colors) +
  scale_size(range = c(2, 12), guide = "none") +
  facet_wrap(~Continent) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 18),
    plot.title = element_text(face = "bold", size = 24),
    axis.title = element_text(size = 18),
    plot.subtitle = element_text(size = 18),
    plot.caption = element_text(size = 14),
    legend.position = "bottom",
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  ) +
  labs(
    title = "GDP per Capita vs Life Expectancy by Continent",
    subtitle = "Year: {frame_time}",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    caption = "Data source: Gapminder",
    color = "Continent"
  ) +
  transition_time(as.integer(year)) +
  ease_aes("linear")
```

After seeing the exponential relationship between Life Expectancy and GDP per Capita, we decided to perform a base ten log transformation on GDP, to more accurately fit the linear model. The plot of Life Expectancy and log<sub>10</sub>(GDP per Capita) is displayed above. 

## 2.2 Linear Regression

```{r}
ggplot(gdp_lex_mean, aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  scale_alpha(range = c(0.2, 0.9), guide = "none") +
  theme_minimal() +
  labs(
    title = "Life Expectancy vs Log GDP per Capita",
    x = expression(log[10]("GDP per Capita")),
    y = "",
    subtitle = "Life Expectancy (years)"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 24),
    axis.title = element_text(size = 18),
    plot.subtitle = element_text(size = 18),
    plot.caption = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14)
  )
```


### Estimated Regression Model

```{r}
gdp_lex_lm <- lm(avg_life_expectancy ~ avg_gdp, data = gdp_lex_mean)

gdp_lex_lm |>
  tidy() |> 
  mutate(term = case_when(term %in% "(Intercept)" ~ "Intercept", 
                          term %in% "avg_gdp" ~ "Slope")) |> 
  gt() |> 
  tab_header(title = md("**Regression Model Estimates**"), 
             subtitle = "Average Life Expectancy vs Average logGDP") |> 
  fmt_number(columns = c(estimate, std.error, statistic, p.value), 
             decimals = 2) |> 
  data_color(columns = p.value, 
             colors = col_numeric(palette = c("green", "yellow", "red"), 
                                  domain = c(0, 0.05, 1))) |> 
  cols_label(estimate = "Estimate", std.error = "Std. Error", 
             statistic = "t-Statistic", p.value = "p-Value", 
             term = "Term") |> 
  tab_style(style = list(cell_fill(color = "lightgray"), 
                         cell_text(weight = "bold")), 
            locations = cells_row_groups()) |> 
  tab_footnote(footnote = "P-values below 0.05 indicate statistical significance.", 
               locations = cells_column_labels(columns = p.value))
```

The estimated regression model: $$ \hat{Life\space Expectancy} = `r round(tidy(gdp_lex_lm)$estimate[1], 2)` + `r round(tidy(gdp_lex_lm)$estimate[2], 2)`*\log_{10}(GDP\space per\space Capita) $$

corresponds to the scatter plot above of Life Expectancy vs. log GDP per Capita. $\hat{Life\space Expectancy}$ represents the predicted life expectancy (in years) and the $log_{10}(GDP\space per\space Capita)$ represents the log-transformed GDP per Capita. When the log-transformed GDP is 0 dollars, the estimated life expectancy is `r round(tidy(gdp_lex_lm)$estimate[1], 2)` years. However, this value is not meaningful since real-world GDP per Capita values are always positive, making a log-transformed value of zero unrealistic. The coefficient of the slope, `r round(tidy(gdp_lex_lm)$estimate[2], 2)`, represents the expected change in life expectancy for a percentage change in GDP per Capita. In other words, a 1% increase in GDP per Capita leads to an increase of approximately `r round(tidy(gdp_lex_lm)$estimate[2]*log10(1.01), 2)` years in life expectancy.

### Estimated Regression Models per Continent

To find the linear regression models for each continent, the average GDP and life expectancy was calculated for each country across all years. This way, each country has just one value for GDP (response variable) and one for life expectancy (explanatory variable).

```{r}
models <- gdp_lex |>
  group_by(Country) |>
  mutate(
    avg_gdp = mean(log_gdp, na.rm = TRUE),
    avg_lex = mean(life_expectancy, na.rm = TRUE)
  ) |>
  distinct(Country, .keep_all = TRUE) |>
  select(-year) |>
  ungroup() |>
  group_by(Continent) |>
  nest() |>
  mutate(
    model = map(data, ~ lm(avg_lex ~ avg_gdp, data = .x)),
    summary = map(model, tidy)
  ) |>
  select(Continent, summary) |>
  unnest(summary) |> 
  mutate(term = case_when(term %in% "(Intercept)" ~ "Intercept", 
                          term %in% "avg_gdp" ~ "Slope"))

models |>
  gt(groupname_col = "Continent") |>
  tab_header(
    title = md("**Regression Model Estimates by Continent**"),
    subtitle = "Average Life Expectancy vs Average GDP"
  ) |>
  fmt_number(
    columns = c(std.error, estimate, statistic),
    decimals = 2
  ) |> 
  fmt_number(
    columns = c(p.value),
    decimals = 3
  ) |>
  cols_label(
    term = "Continent",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "t-Statistic",
    p.value = "p-Value"
  ) |>
  tab_spanner(
    label = "Model Estimates",
    columns = c(estimate, std.error, statistic, p.value)
  ) |>
  data_color(
    columns = p.value,
    colors = col_numeric(
      palette = c("green", "yellow", "red"),
      domain = c(0, 0.05, 1)
    )
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "lightgray"),
      cell_text(weight = "bold")
    ),
    locations = cells_row_groups()
  ) |>
  tab_footnote(
    footnote = "P-values below 0.05 indicate statistical significance.",
    locations = cells_column_labels(columns = p.value)
  )
```

::: panel-tabset
#### Asia

```{r}
gdp_lex_mean |> 
  filter(Continent == "Asia") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "Asia",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[1], 2)` + `r round(models$estimate[2], 2)`*log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[2], 2)` represents a `r round(models$estimate[2]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[2], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[1], 2)` represents the predicted life expectancy. 

#### Africa

```{r}
gdp_lex_mean |> 
  filter(Continent == "Africa") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "Africa",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[3], 2)` + `r round(models$estimate[4], 2)`*\log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[4], 2)` represents a `r round(models$estimate[4]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[4], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[3], 2)` represents the predicted life expectancy. 

#### Europe

```{r}
gdp_lex_mean |> 
  filter(Continent == "Europe") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "Europe",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[5], 2)` + `r round(models$estimate[6], 2)`*\log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[6], 2)` represents a `r round(models$estimate[6]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[6], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[5], 2)` represents the predicted life expectancy. 

#### South America

```{r}
gdp_lex_mean |> 
  filter(Continent == "South America") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "South America",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[7], 2)` + `r round(models$estimate[8], 2)`*\log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[8], 2)` represents a `r round(models$estimate[8]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[8], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[7], 2)` represents the predicted life expectancy. 

#### North America

```{r}
gdp_lex_mean |> 
  filter(Continent == "North America") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "North America",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[9], 2)` + `r round(models$estimate[10], 2)`*\log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[10], 2)` represents a `r round(models$estimate[10]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[10], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[9], 2)` represents the predicted life expectancy. 

#### Oceania

```{r}
gdp_lex_mean |> 
  filter(Continent == "Oceania") |> 
  ggplot(aes(x = avg_gdp, y = avg_life_expectancy)) +
  scale_color_manual(values = continent_colors) +
  geom_point(aes(colour = Continent)) +
  geom_smooth(method = "lm", colour = "black") +
  theme_minimal() +
  labs(
    title = "log GDP per Capita vs Life Expectancy",
    subtitle = "Oceania",
    x = expression(log[10]("GDP per Capita")),
    y = "Life Expectancy (years)")
```

$$ \hat{Life\space Expectancy} = `r round(models$estimate[11], 2)` + `r round(models$estimate[12], 2)`*\log_{10}(GDP\space per\space Capita) $$

The slope coefficient of `r round(models$estimate[12], 2)` represents a `r round(models$estimate[12]*log10(1.01), 4)` year increase in predicted life expectancy for a one percent increase in GDP per Capita. Alternatively a doubling GDP per Capita is associated with a `r round(models$estimate[12], 2)` year increase in predicted life expectancy. 

Although unrealistic, when log<sub>10</sub>(GDP per Capita) is equal to zero, the intercept of `r round(models$estimate[11], 2)` represents the predicted life expectancy. 
:::

## 2.3 Model Fit

```{r}
# The variance in the response values.
res_var <- var(augment(gdp_lex_lm)$avg_life_expectancy)

# variance in the fitted values from your regression model
fitted_var <- var(augment(gdp_lex_lm)$.fitted)

# The variance in the residuals from your regression model.
resid_var <- var(augment(gdp_lex_lm)$.resid)

# Calculate R-squared and proportion of variability explained
r_squared <- fitted_var / res_var

# create table
tibble(
  "Response" = round(res_var, 2),
  "Fitted Values" = round(fitted_var, 2),
  "Residuals" = round(resid_var, 2),
  "R²" = round(r_squared, 4)
) |>
  kable(caption = "Variances and Model Fit")

```

Our table demonstrates that the model provides a moderate fit to the data. Our proportion of variance, or R² value of `r round(r_squared, 4)`, indicates about `r round(r_squared, 4)*100`% of the total variation in life expectancy across countries is explained by our GDP based model. This suggests that GDP plays a role in determining life expectancy, but is not the sole factor. The remaining `r 100 - round(r_squared, 4)*100`% of the variation must be explained by other factors, such as healthcare spending, education levels, and environmental factors.

# 3 Simulation

## 3.1 Visualizing Simulations from the Model

```{r}
# predict function and random errors
gdplex_pred <- predict(gdp_lex_lm)
  
gdplex_sig <- sigma(gdp_lex_lm)

simulated_data = tibble(simulated_life_expectancy = gdplex_pred + rnorm(length(gdplex_pred), mean = 0, sd = gdplex_sig))

full_data <- gdp_lex_mean |>
  bind_cols(simulated_data) |>
  group_by(Country) |>
  summarize(avg_simulated_life_expectancy = mean(simulated_life_expectancy, na.rm = TRUE)) |>
  left_join(gdp_lex_mean, join_by(Country == Country))

```

```{r}
#| fig-width: 12
#| fig.height: 8

# Create side-by-side plots
p1 <- ggplot(full_data, aes(x = avg_gdp, y = avg_life_expectancy, color = Continent)) +
  geom_point() +
  geom_smooth(method = "lm", color = "black") +
  scale_color_manual(values = continent_colors) +
  theme_minimal() +
  labs(title = "Observed Data: GDP vs Life Expectancy",
      x = expression(log[10]("GDP per Capita")),
      y = "",
      subtitle = "Life Expectancy (years)") +
    theme(
    plot.title = element_text(face = "bold", size = 20),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

p2 <- ggplot(full_data, aes(x = avg_gdp, y = avg_simulated_life_expectancy, color = Continent)) +
  geom_point() +
  geom_smooth(method = "lm", color = "black") +
  scale_color_manual(values = continent_colors) +
  theme_minimal() +
  labs(title = "Simulated Data: GDP vs Life Expectancy",
      x = expression(log[10]("GDP per Capita")),
      y = "",
      subtitle = "Simulated Life Expectancy (years)") +
  theme(
    plot.title = element_text(face = "bold", size = 20),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )

p1 + p2
```

From the graphs above, it can be seem that both the observed and the simulated data have a strong, positive, and linear relationship between life expectancy and log<sub>10</sub>(GDP per Capita). This trend suggests that as GDP per Capita increases, life expectancy also increases. The black, fitted regression line and grey shaded confidence interval are similar in both plots indicating that both data sets have similar variability. In both plots the distribution of points by continent are consistent with lower income continents, such as Africa, clustering around the bottom of the line. This is the same with wealthier continents such as Europe and North America, which appear at the higher end of the trend line in both graphs.  

Although they the observed and simulated plots are similar, one difference is the spread of the points. The simulated data appears to have more constrained spread, with points falling fairly symmetrically on either sides of the trend line. This suggests that the model was not able to fully capture the variability present in the observed data. It can also be seen that the observed data contains outlines while the simulated data only replicates the main trend, but not the extreme observations. 

## 3.2 Generating Multiple Predictive Checks

```{r}
sim_and_calc_rsquared <- function(x) {
  predicted_values <- predict(gdp_lex_lm)
  
  # Generate simulated values based on our model
  sim_life_exp <- predicted_values + 
                  rnorm(length(predicted_values), 
                        mean = 0, 
                        sd = sigma(gdp_lex_lm))

  # Create dataset with observed and simulated values
  sim_data <- tibble(
    observed_life_expectancy = gdp_lex_mean$avg_life_expectancy,
    simulated_life_expectancy = sim_life_exp,
    avg_gdp = gdp_lex_mean$avg_gdp
  ) |>
  drop_na()

  # Regress the observed values against simulated ones
  sim_model <- lm(observed_life_expectancy ~ simulated_life_expectancy, data = sim_data)


  r_squared <- glance(sim_model)$r.squared

  return(r_squared)
}

r2_values <- map_dbl(1:1000, sim_and_calc_rsquared)

sim_results <- enframe(r2_values, name = NULL, value = "r2")

# Visualization of R² distribution
ggplot(sim_results, aes(x = r2)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  # our actual r_squared value from our original lm model
  geom_vline(xintercept = r_squared, color = "red", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of R² Values from 1000 Simulated Datasets",
    x = "R² Value",
    y = "",
    subtitle = "Frequency"
  )


```

The shape of the R² distribution is roughly normal, with a mean of `r round(mean(r2_values), 2)`, median of `r round(median(r2_values), 2)` and a standard deviation of `r round(sd(r2_values), 4)`. The observed R² (red dashed line) of `r round(r_squared, 4)` from our actual data is higher than all of the simulated values. This suggests that the true relationship in the observed data is stronger than what the simulation produces. In other words the model is only a decent representation of the actual data since it underestimates the strength of the relationship between life expectancy and log<sub>10</sub>(GDP per Capita). 

# Conclusion

Throughout this report we sought out to seek an understanding on how economic factors on health, such as GDP per Capita can be used as a predictor of life expectancy.

Based on our findings we are able to conclude that in general, there is a moderate, positive relationship between GDP per Capita and life expectancy. Meaning as the GDP per Capita increases, the life expectancy of a country, and the overall continent is also expected to increase. Our linear regression model shows that after log transforming GDP per Capita, a one percent increase in GDP leads to an increase of `r round(tidy(gdp_lex_lm)$estimate[2]*log10(1.01), 2)` years in life expectancy. 

With our predicted model fit, `r round(r_squared, 4)*100` percent of the total variance can be attributed by GDP per Capita, meaning that GDP is not the sole factor of increasing life expectancy. One other factor can be attributed to region. Looking back at the individual continent's regression model,  Europe and North America have the highest slopes, then followed by Oceania. This makes sense as these continents have a higher density of wealthy countries with better infrastructure and more access to different industries compared to the other remaining, which have a lot higher variability.

Our simulation model shows a similar trend to our observed, a moderately positive relationship, with Europe and North America at the higher ends of the regression line. One key difference is that the spread is more constrained in the simulated data compared to our observed.